---
title: "SWPPredictor_RandomForest"
author: "Giriraj"
date: "2022-11-30"
output:
  html_document:
    df_print: paged
---

```{r}
# importing dataset for the prediction model
library(readr)
df <- read_csv("dataset/dataset.csv")
head(df)
```

```{r}
# manipulating data to represent actual value
# multiplying variable SWP * 100 to represent SWP received per month
# multiplying variable Annual_income_and_SWP * 1000 to represent Annual_income_and_SWP received per year

df$Annual_income_and_SWP <- df$Annual_income_and_SWP *1000
df$SWP <- df$SWP * 100
df
```

```{r}
# mutating data for better analysis and reporting

library(dplyr)
df%>%
  mutate(Education=case_when(
    .$Education=="High School or GED" ~ 1,
    .$Education=="Some highschool" ~ 1,
    .$Education=="Completed Undergraduate" ~ 2,
    .$Education=="Some Undergraduate" ~ 2,
    .$Education=="Completed Masters" ~ 3,
    .$Education=="Some Maters" ~ 3,
    .$Education=="Some Phd" ~ 4,
    .$Education=="Completed Phd" ~ 4
    )) -> df
```

```{r}
# mutating data for better analysis and reporting

df%>%
  mutate(Age=case_when(
    .$Age =="18-29" ~ 1,
    .$Age =="30-44" ~ 2,
    .$Age =="45-60" ~ 3,
    .$Age =="Greater than 60" ~ 4
    )) -> df
```

```{r}
# mutating data for better analysis and reporting

df%>%
  mutate(Household_Income=case_when(
    .$Household_Income =="$0-$9,999" ~ 1,
    .$Household_Income =="$10,000-$24,999" ~ 2,
    .$Household_Income =="$25,000-$49,999" ~ 3,
    .$Household_Income =="$50,000-$74,999" ~ 4,
    .$Household_Income =="$75,000-$99,999" ~ 5,
    .$Household_Income =="$100,000-$124,999" ~ 6,
    .$Household_Income =="$125,000-$149,999" ~ 7,
    .$Household_Income =="$150,000-$174,999" ~ 8,
    .$Household_Income =="$175,000-$199,999" ~ 9,
    .$Household_Income =="$200,000+" ~ 10,
    .$Household_Income =="Prefer not to answer" ~ 996
    )) -> df
```

```{r}
head(df)
```

```{r}
# filtering individuals who receive SWP for histogram
df$SWP_valid <- df$SWP != 0

# generating histogram for SWP to view its distribution
hist(df$SWP[df$SWP_valid], breaks = 20, main = 'Histogram for people who receive SWP', xlab = 'SWP amount (US dollar)')
```

```{r}
# generating scatter plot for SWP and Annual Income to view its relationship and understand its dependancy
plot(x = df$SWP[df$SWP_valid], y = df$Annual_income_and_SWP[df$SWP_valid], main = 'Scatter plot for SWP & Annual income', xlab = 'SWP', ylab = 'Annual income')

# printing mean and standard deviation for SWP
print(paste('Mean value for SWP =', round(mean(df$SWP[df$SWP_valid]), 2)))
print(paste('Mean value for SWP =', round(sd(df$SWP[df$SWP_valid]), 2)))
```

**Creating linear regression models to get best set of variables to predict SWP**

```{r}
# linear regression model for SWP ~ .
m1 <- lm(SWP ~. , data = df)
summary(m1)
```

```{r}
# trimming features & checking for Multiple R-squared & Adjusted R-squared values to determine best suited features for decision tree
m2 <- lm(SWP ~ Education + Own_computer + days_hospitalized + Disabled + Internet_access + Live_with_parents + Length_of_resume_gap_month + Annual_income_and_SWP + Unemployed + Read_books + Times_hospitalized + Age + Household_Income,data = df)
summary(m2)
```

```{r}
# trimming features & checking for Multiple R-squared & Adjusted R-squared values to determine best suited features for decision tree
m3 <- lm(SWP ~ Education + days_hospitalized + Disabled + Internet_access + Live_with_parents + Length_of_resume_gap_month + Annual_income_and_SWP + Unemployed + Read_books + Times_hospitalized + Age + Household_Income, data = df)
summary(m3)
```

```{r}
# trimming features & checking for Multiple R-squared & Adjusted R-squared values to determine best suited features for decision tree
m3.1 <- lm(SWP ~ Education + days_hospitalized + Disabled + Internet_access + Live_with_parents + Length_of_resume_gap_month + Annual_income_and_SWP + Unemployed + Read_books + Age + Household_Income,data = df)
summary(m3.1)
```

```{r}
# trimming features & checking for Multiple R-squared & Adjusted R-squared values to determine best suited features for decision tree
m4 <- lm(SWP ~ Education + Disabled + Internet_access + Live_with_parents + Length_of_resume_gap_month + Annual_income_and_SWP + Unemployed + Read_books + Times_hospitalized + Age + Household_Income,data = df)
summary(m4)
```

```{r}
# trimming features & checking for Multiple R-squared & Adjusted R-squared values to determine best suited features for decision tree
m5 <- lm(SWP ~ Education + Disabled + Internet_access + Length_of_resume_gap_month + Annual_income_and_SWP + Unemployed + Read_books + Times_hospitalized + Age + Household_Income,data = df)
summary(m5)
```

```{r}
# trimming features & checking for Multiple R-squared & Adjusted R-squared values to determine best suited features for decision tree
m6 <- lm(SWP ~ Education + Disabled + Internet_access + Annual_income_and_SWP + Unemployed + Read_books + Times_hospitalized + Age + Household_Income, data = df)
summary(m6)
```


**Comparing Multiple R-squared and Adjusted R-squared of different Linear Regression Models to select the best set of variables**
\

```{r}
# creating new dataset, m_data with the features in focus gathered from model 6 of linear regression
m_data <- df %>% select(2, 5, 6, 9, 10, 11, 12, 13, 14, 15) 
head(m_data)
```
\
**Performing Logistic Regression to calculate accuracy of variables being able to predict SWP**
\
```{r}
# creating new dataset, log_data for logistic regression
log_data <- m_data
```

```{r}
# mutatting values of SWP to represent if the person receives money from SWP or not  
log_data$SWP[log_data$SWP != 0] <- 1
head(log_data)
```

```{r}
library(caTools)
library(ROCR)
```

```{r}
set.seed(10086)

# splitting dataset with split-ratio of 80%-20%
split <- sample.split(log_data$SWP, SplitRatio = 0.8)
train_reg <- subset(log_data, split == "TRUE")
test_reg <- subset(log_data, split == "FALSE")
```

```{r}
# training model
logistic_model <- glm(SWP ~ Education + Disabled + Internet_access + Annual_income_and_SWP + Read_books + Times_hospitalized + Household_Income, data = train_reg, family = "binomial")

# summary
summary(logistic_model)
```

```{r}
# predicting train and test data based on model
predict_reg_train <- predict(logistic_model, train_reg, type = "response")
predict_reg <- predict(logistic_model, test_reg, type = "response")
```

```{r}
# calculating probabilities
predict_reg_train <- ifelse(predict_reg_train >0.5, 1, 0)
predict_reg <- ifelse(predict_reg >0.5, 1, 0)
```

```{r}
# evaluating model accuracy using confusion matrix

table(train_reg$SWP, predict_reg_train)
table(test_reg$SWP, predict_reg)
```

```{r}
# generating accuracy

missing_classerr_train <- mean(predict_reg_train != train_reg$SWP)
acc_train <- round(1 - missing_classerr_train, 2)
print(paste('Accuracy for Train data =', acc_train))

missing_classerr <- mean(predict_reg != test_reg$SWP)
acc_test <- round(1 - missing_classerr, 2)
print(paste('Accuracy for Test data =', acc_test))
```

```{r}
set.seed(10086)
# ROC-AUC curve
ROCPred <- prediction(predict_reg, test_reg$SWP)
ROCPer <- performance(ROCPred, measure = "tpr",
							x.measure = "fpr")

auc <- performance(ROCPred, measure = "auc")
auc <- auc@y.values[[1]]
auc
```

```{r}
# plotting curve
plot(ROCPer)
plot(ROCPer, colorize = TRUE,
	print.cutoffs.at = seq(0.1, by = 0.1),
	main = "ROC CURVE")
abline(a = 0, b = 1)

auc <- round(auc, 4)
legend(.8, .2, auc, title = "AUC", cex = 1)
```

\
**Generating Decision Tree for variables selected from Linear Regression Model 6**
\
```{r}
library(DAAG)
library(party)
library(rpart)
library(rpart.plot)
library(mlbench)
library(caret)
library(pROC)
library(tree)
```

```{r}
# subsetting data in ratio 80%-20% for decision tree
set.seed(1234)
ind <- sample(2, nrow(m_data), replace = T, prob = c(0.6, 0.4))
train <- m_data[ind == 1,]
test <- m_data[ind == 2,]

# generating and plotting tree
tree <- rpart(SWP ~., data = train)
rpart.plot(tree)
```

\
**Constructing Random Forest model for predicting SWP using variables selected from Linear Regression Model 6**
\

```{r}
# creating prediction model and constructing its confusion matrix

library(randomForest)
library(caret)
library(e1071)

set.seed(10086)

model <- randomForest(formula = SWP ~ ., data = m_data)
model

which.min(model$mse)
sqrt(model$mse[which.min(model$mse)])
plot(model)
varImpPlot(model)

# constructing random forest
predictor_variable <- select(m_data, -SWP)
model_tuned <- tuneRF(
               x=predictor_variable, # defining predictor variables
               y=m_data$SWP, # defining response variable
               ntreeTry=500, # number of tree Random Forest will construct
               mtryStart=4, 
               stepFactor=1.5,
               improve=0.01,
               trace=FALSE #don't show real-time progress
               )

# splitting data into train and test subsets
set.seed(10086)
ind <- sample(2, nrow(m_data), replace = T, prob = c(0.8, 0.2))
train <- m_data[ind == 1,]
test <- m_data[ind == 2,]

# Random Forest for train data
classifier_RF = randomForest(x = train[-7],
                             y = train$SWP,
                             ntree = 500)

classifier_RF

y_pred = predict(classifier_RF, newdata = test[-7])

# confusion matrix
confusion_mtx = table(test$SWP, y_pred)
confusion_mtx

plot(classifier_RF)

# importance plot
importance(classifier_RF)

# cariable importance plot
varImpPlot(classifier_RF)
```

```{r}
# creating new data to predict SWP value
new <- data.frame(Education=2, Disabled=0, Internet_access=1, Annual_income_and_SWP=58000, Unemployed=0, Read_books=1, Times_hospitalized=1, Age=4, Household_Income=6)

new1 <- data.frame(Education=4, Disabled=1, Internet_access=1, Annual_income_and_SWP=58000, Unemployed=1, Read_books=0, Times_hospitalized=1, Age=4, Household_Income=5)

new2 <- data.frame(Education=2, Disabled=0, Internet_access=1, Annual_income_and_SWP=100000, Unemployed=0, Read_books=1, Times_hospitalized=0, Age=2, Household_Income=8)

new3 <- data.frame(Education=2, Disabled=1, Internet_access=0, Annual_income_and_SWP=1000, Unemployed=1, Read_books=0, Times_hospitalized=20, Age=3, Household_Income=4)

new4 <- data.frame(Education=4, Disabled=0, Internet_access=1, Annual_income_and_SWP=100000, Unemployed=0, Read_books=1, Times_hospitalized=0, Age=2, Household_Income=10)


predict(model, newdata=new)
predict(model, newdata=new1)
predict(model, newdata=new2)
predict(model, newdata=new3)
predict(model, newdata=new4)
```

